Vagueness \& Rationality

### 2018-10-27 manuscript/v-book-2018.tex
was chapter_latex
removed undirected quote marks and replaced with aostrophe-style quotes
merged separate tex into v-book-2018.tex

\documentclass[ %
  graybox       %
 ,envcountchap  %
 ,sectrefs      %
 ,footinfo      %
 ,graphics      %
%,referee
]{svmono}
% Native packages
\usepackage{mathptmx}
\usepackage{helvet}
\usepackage{courier}
\usepackage{type1cm}
\usepackage{makeidx}		    % allows index generation
\usepackage{graphicx}		    % standard LaTeX graphics tool
\usepackage{multicol} 		    % used for the two-column index
\usepackage[bottom]{footmisc}	% places footnotes at page bottom
% User packages
\usepackage{natbib}
\usepackage{lipsum}
\usepackage{placeins}
\usepackage{multirow}			% added by Matt
\usepackage{marginnote}			% added by Matt
%\makeindex             		% used for the subject index: please use the style svind.ist with your makeindex program
\begin{document}
\author{Kees van Deemter and Matthew Green}
\title{Vagueness \& Rationality}
\subtitle{The Elusive Benefits of Vagueness: the Evidence So Far}
\maketitle
\frontmatter
\tableofcontents
\mainmatter
\chapter{The Elusive Benefits of Vagueness: Evidence from Experiments}\label{chapterlabel}
\input{abstract}
\section{Introduction}\label{introduction}\input{introduction}
\section{Related work}\label{related-work}\input{related-work}
\section{Our Approach to the Problem}\label{our-approach-to-the-problem}\input{our-approach-to-the-problem}
\section{Experiment 1: separating vagueness from instruction format}\label{e1}\input{e1}
\section{Experiment 2: focus on instructions that contain numerals}\label{e2}\input{e2}
\section{Experiment 3: focus on instructions that do not contain numerals}\label{e3}\input{e3}
\section{Discussion of experiments 2 and 3}\label{discussion-of-e2-and-e3}\input{discussion-of-e2-and-e3}
\section{General Discussion}\label{general-discussion}\input{general-discussion}
\backmatter
\bibliographystyle{spbasic}
\bibliography{v-book-2018}
\backmatter
%\printindex
\end{document}

>>>>

\documentclass[ %
  graybox       %
 ,envcountchap  %
 ,sectrefs      %
% ,footinfo     %
% ,referee
]{svmono}
\usepackage{mathptmx}
\usepackage{helvet}
\usepackage{courier}
\usepackage{type1cm}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage[bottom]{footmisc}
\usepackage{natbib}
\usepackage{lipsum}
\usepackage{placeins}
\usepackage{multirow}
\begin{document}
\author{Matthew James Green and Kees van Deemter}
\title{Vagueness \& Rationality}
\subtitle{The Elusive Benefits of Vagueness: The Evidence So Far}
\date{}
\maketitle
\frontmatter
\tableofcontents
\mainmatter

\chapter{The Elusive Benefits of Vagueness: Evidence from Experiments}
Much of everyday language is vague, even in situations where vagueness could have been avoided (i.e., where vagueness is used ``strategically'').
Yet the benefits of vagueness for hearers and readers are proving to be elusive.
We discuss a range of earlier controlled experiments with human participants, and we report on a new series of experiments that we ourselves have conducted in recent years.
These experiments, which focus on vague expressions that are part of referential noun phrases, aim to separate the utility of vagueness (as defined by the existence of borderline cases) from the utility of other factors that tend to co-occur with vagueness.
After presenting the evidence, we argue that it supports a view where the benefits that vague terms exert are due to other influences, and not to vagueness itself.

\section{Introduction}
In academic use, the word `vagueness' has a specific meaning.
Keefe and Smith, for example, state that ``vague predicates have borderline cases, have fuzzy boundaries, and are susceptible to sorites paradoxes'' \citep[p.~4]{keefe1997vagueness}, as do \citet{EgreKlinedinst}.
The crucial criterion is the existence of borderline cases: ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise'' \citet[p.~1]{lipmanvague}.
A typical example is the word ``tall'', as applied to people for example, because there is no precise, known height which separates those who are tall from those who are not.
The crucial point is that ``tall'' admits borderline cases (i.e., people who may or may not count as tall), which are the hallmark of vagueness as we use the term.

Linguists, philosophers of language, and more recently game theorists have asked why natural languages contain so many vague expressions \citet{Lipman:2000fk, lipmanvague}, which are
often used even in situations where the speaker could have used an expression that is not vague (i.e., crisp); in these situations we say that vagueness is used \emph{strategically}.
By introducing borderline cases, these expressions create potential misunderstandings, thereby creating ``a worldwide several-thousand year efficiency loss'' \citet[][p.~1]{lipmanvague}.
Lipman explains the point by means of a scenario in which a speaker describes a person to a hearer, who needs to identify that person in the arrivals hall of of an airport.
In such a scenario, a precise description of the person's height (e.g., ``The person's height is 187.96 cm'') would be more useful than a vague one (``The person is tall'').
Lipman uses this scenario to explain why standard game theory models of communication \citep[e.g.,][]{Crawford:1982lr} predict that, under certain conditions, a crisp act of communication will always have more utility than a vague act that communicates the same state of affairs.

Lipman argues that the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there were advantages as well as disadvantages associated with vague expressions.
Lipman asks, essentially, what these advantages might be.
Several tentative answers to Lipman's question have been offered \citep[see][]{van2009utility, vanDeemterBook}.
One of the most promising answers appears to be the idea that vague expressions are easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \citep[e.g.,][]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}.
For example, \citeauthor{lipmanvague} writes: ``For the listener, information which is too specific may require more effort to analyze'' (\citeyear[][p.~11]{lipmanvague}).
We shall refer to this as the \emph{cost reduction} hypothesis.

This article brings an experimental approach to these issues, focussing on vagueness in descriptions (e.g., ``the square with few dots'') and its effect on the hearer's ability to act on a given description, as measured by the time that it takes hearers to click on the referent of a description.\footnote{Other metrics could have been chosen, such as hearers' ability to remember information, for example, or error rates. Although error rates play a minor role in the present paper, for reasons that will become clear, we focus on response times in particular.}
We find that, although hearer benefits from vague descriptions are straightforward to demonstrate in many cases, a closer experimental analysis militates against the conclusion that vagueness itself --- as defined above, in terms of the existence of borderline cases --- lies at the heart of these results.
Instead, it is other factors, such as the presence of an overt numerical expression in the description, that proved to be decisive.
We believe that, despite the fact that our experiments are unavoidably focussed on a specific class of vague expressions (since any experiment can only deal with a limited number of different stimuli), these findings are potentially important, because they call into question whether ``strategic'' vagueness (i.e., vagueness where the speaker had a choice, because she could have produced a crisp expression instead) has any advantages at all.
In other words, returning to Lipman's question, it is possible that vagueness has evolved partly as a necessary evil (e.g., because of the limits of observation and prediction) and partly as a side effect of other factors.

\section{Related work}\label{related-work}
A number of answers to Lipman's question have been proposed, but as we have argued elsewhere, it appears that some of these do not to hold up to scrutiny \citep{van2009utility,vanDeemterBook}. For example,
\begin{itemize}
\item It has been proposed that the existence of vague words makes a language more efficient \citep{barwiseperry}. The idea is that vagueness allows the use of one and the same word (e.g., ``big'') in different situations, which require different standards. For instance, the vague word ``big'' can denote a mouse as well as an elephant. The idea is plausible, but the problem is that ``big'' is efficient not because it is vague (i.e., not because it allows borderline cases) but because it is context dependent. To see this, note for example that superlatives are not vague (i.e., they do not permit borderline cases), but they {\em are} context dependent: ``the biggest $x$'' ascribes different sizes to the referent depending on $x$. It is for this reason that we can use the word ``biggest''  to talk about the biggest mouse, the biggest elephant, and so on. Context-dependence, however, does not imply vagueness.
\item Vague words are capable of combining a statement of quantity with an evaluative statement \citep{veltman}. The idea is that when we say ``The patient has a fever,'' we do not merely assert that her temperature lies above a certain value, we also imply that the deviation is clinically significant (i.e., something is not right with the patient). The problem with this plausible idea is that evaluation does not imply vagueness. For example, the medical term ``obese'' is evaluative (i.e., it is not healthy to be obese), yet its standard medical definition in terms of Body Mass Index is perfectly crisp (i.e., without borderline cases), with all and only BMI values above 30 counting as obese.
\end{itemize}
Analogous observations can be made about a number of other purported benefits of vagueness; one of the few hypotheses left standing at the moment is the {\em cost reduction} hypothesis, which we mentioned in the previous section. The findings on which we are reporting in the present paper will follow a familiar pattern: we will explore the cost reduction hypothesis, only to conclude that it may not stand up to scrutiny. Our approach, this time, will be experimental: we conduct controlled experiments to investigate the effect that vague expressions have on a hearer.

Charting the utility of vagueness is also the attested aim of a small number of previous experimental studies. But, as it happens, few of these studies have truly focussed on vagueness in the sense in which we do here (i.e., they do not address Lipman's challenge). Two recent studies illustrate this issue; let's discuss them briefly.

In a series of studies of behaviour modification, \citet{Mishra01042011} manipulated the presentation format of information about quantities in the domains of mental acuity, physical strength, and weight loss. In the weight loss study, participants were told that the study was designed to test the validity of a new (actually fictitious) health index, the HHI (Holistic Health Index). They were told that an ideal HHI score lies in the range of 45 to 55. In a longitudinal study, participants submitted their weight to a computer each week. Participants were told that two algorithms would be used to compute their HHI, and that the two might give different values initially, in which case the true score lay between the two values. In one condition, which the authors called the precise condition, the two algorithms gave the same score. In the other condition, which the authors called the vague condition, one algorithm added 3\% to the score while the other algorithm subtracted 3\% from the score, yielding a range of values whose midpoint was the same as the two values given in the precise condition.

One group of participants was given HHI scores in the ideal range: for this group their weight loss did not differ depending on whether they were given vague or precise HHI values. However for the other group, who were given HHI scores outside the ideal range, their weight loss was significantly greater if they were given vague HHI scores than if they were given precise HHI scores. The authors explain the improvement in the vague condition for this group as resulting from the participants' freedom to think of themselves a positioned on one end of the range --- the end closest to the ideal HHI scores. This ``illusion of proximity'' \citep[][p.~4]{Mishra01042011} to the goal is argued to have allowed participants to generate positive expectancies that lead to behaviours that improved performance. In contrast, in the precise conditions, participants did not have this freedom of interpretation, and could not distort the information to bring about the beneficial \emph{illusion of proximity}. These results are interesting, and of obvious potential practical importance. We note, however, that information presented as an exact range of values does not conform with the standard definition of vagueness \citep{keefe1997vagueness, EgreKlinedinst}, since an exact range does not admit borderline cases. In the terminology of \citeauthor{Hobbs85granularity}, the difference between a range and a single midpoint value is a difference of \emph{granularity}. Furthermore, the experiments of \citeauthor{Mishra01042011} did not explore benefits in terms of processing cost, but in terms of long-term behaviour change.

Similar issues arise from the work of \citet{peters2009bringing}. The authors carried out a series of studies where participants were required to rate hospitals based on various sources of information about quality of care. There was a between-subjects manipulation based on numeracy. The format of the information was manipulated within subjects: either numbers only were presented, or both numbers and evaluative categories were presented (e.g., \emph{Poor}, \emph{Fair}, \emph{Good}, \emph{Excellent}, with crisp visual boundary lines between the categories). Results showed that, for low-numeracy participants, the presence of evaluative categories resulted in a diminished influence of an irrelevant affective state on the ratings. For all participants, the presence of evaluative categories resulted in better decisions and in a greater use of the most important and reliable types of information, such as survival rates.

It is, however, questionable whether the ``evaluative categories'' manipulation in this study can be considered a manipulation of vagueness. Certainly, terms like \emph{Fair} admit the possibility of borderline cases. However, given that the boundaries between the categories were marked crisply, and that therefore the categories mapped crisply to numerical values, it becomes doubtful whether any borderline cases could be conceived to arise in fact. For example, \emph{Fair} was mapped to 60--70\% for the variable \emph{percentage of heart attack patients given recommended treatment (ACE inhibitor)}. Accordingly, rather than the vagueness of categories such as \emph{Poor}, Peters et al. emphasise the evaluative content inherent in these categories, and the affective potential of the evaluative content rather than the vagueness of the terms like {\em Fair}.

\section{Our Approach to the Problem}
Are vague expressions processed more easily by readers than crisp ones? Like Lipman, we focus on situations where numerical information is used in order to identify a referent. Reference, in other words, will be the communicative task on which we focus, partly because of the interest that this topic has recently drawn in various areas of Cognitive Science \citep{vanDeemterCMR}. By looking at one specific type of vagueness, we will be able to investigate the costs and benefits of vagueness relatively thoroughly. Whether our findings generalise to other uses of vagueness is a question on which we will speculate in the final section of this chapter.

We have chosen a narrative strategy in which we address a sequence of four experiments with human readers chronologically, explaining how each experiment helped us refine our research question. In order to do justice to our findings, we need to describe these experiments in a fair amount of detail.

Let us start by explaining the task that was given to the participants in our experiments. We used a {\em speeded forced choice} task to compare the processing costs of different references to quantities. In this context, speed and accuracy of responses are the key dimensions on which the different references can be compared. Each stimulus in the experiments was a set of dot arrays containing various number of dots, together with a preceding instruction (in the form of a referring expression) to choose one of the arrays with respect to its cardinality. The participant was asked to respond as quickly as possible while avoiding errors. We manipulated the instructions and the arrays in several ways across the four experiments.

All the experiments shared the following properties: stimuli were created using the language GNU Octave \citep{eaton:2002} and the Psychophysics Toolbox extensions \citep{ptbx1, ptbx2}. The position of the dots was randomised per-trial. The order in which trials were presented was randomised per-participant. There were 256 trials, presented in 4 blocks of 64 each, between which the participant could rest. A MacBook Pro laptop computer with a 13-inch screen presented the stimuli to the participants and recorded responses. Participants were recruited using email lists at the University of Aberdeen, and paid ten pounds for participating. All participants self-reported fluency in English, and had normal, or corrected-to-normal vision. The experiment was conducted in a quiet room. Participants were asked to respond as quickly as possible while avoiding errors. There was a block of practice trials after which participants could ask questions, following which the experimenter left the room. All $p$ values reported for linear models were calculated using the R package \emph{lmerTest} \citep{lmerTest}.

When the distance grows between two numbers, they become more easily distinguishable: the \emph{numerical distance effect} has been shown for comparing the cardinality of two sets of dots \citep{van123} and for processing Arabic numerals and number words \citep{Dehaene1996}. We manipulated the number of dots in each array such that some sets of arrays had smaller numerical distances and others had larger numerical distances. Where a number was mentioned in the instructions, it was always in the form of an Arabic numeral. When two numbers are presented with the smaller on the left, this left-side presentation facilitates responses indicating the smaller number: the \emph{Spatial-Numerical Association of Response Codes (SNARC)} effect \citep{dehaene1993mental, gevers2006automatic}. We controlled which side the smaller number appeared on to avoid systematic influences of this effect.

There is abundant evidence \citep[e.g.,][]{trick1994small} that very small (i.e., \emph{subitizable}) quantities are recognised and processed by a distinct psychological mechanism that differs from that used to process larger quantities. We performed a pilot experiment \citep{green2011costreduction} in which we were able to confirm this finding in the experimental settings on which we are focussing in this paper. We found that, when participants were confronted with a stimulus consisting of two squares containing different numbers of dots.\footnote{Such a stimulus is referred to hereafter as consisting of a set of {\em dot arrays} The number of dots in an array is referred to as its cardinality. The physical arrangement of dots in each array is irregular.}, instructions of the form \emph{Choose the square with \emph{n} dots} led to consistently faster response times than instructions of the form \emph{Choose the square with many/few dots} when $2 \leq n \leq 5$; the converse was true for $n>5$. Given these findings, we henceforth focussed our studies on non-subitizable numbers, because it is there that vagueness is expected to have benefits.

In a second pilot experiment \citep{green2013utility}, we again presented two dot arrays with the instruction in the form \emph{Choose the square with \ldots dots}. The arrays contained larger numbers of dots than in the first pilot: one array always contained 25 dots, and the other contained either 5, 10, 15, 20, 30, 35, 40, or 45 dots. Each stimulus can therefore be seen in terms of the numerical difference between the number of dots in one array and the number of dots in the other: giving numerical distances of 5, 10, 15, and 20, with smaller numerical distances resulting in less discriminable arrays and larger distances resulting in more discriminable arrays. Our main manipulation was of the vagueness of the instruction, with two levels, \emph{crisp} and \emph{vague}. Assuming the dot array (5, 25), and the instruction referring to the smaller cardinality, the \emph{crisp} instruction was \emph{Choose the square with 5 dots} and its \emph{vague} counterpart was \emph{Choose the square with few dots}. We found, as expected, that responses were faster and more accurate for vague instructions than for crisp instructions. We also found an interaction between vagueness and numerical distance such that there were diminishing returns for vagueness as numerical distance increased, until, at the biggest numerical distance, there was no real difference any more between crisp and vague instructions. We interpreted this pattern as showing that cognitive load is relatively easy in both conditions when the stimuli are most discriminable, and so vagueness confers no additional advantage in those most easily discriminable stimuli.

However, the picture painted by these findings from the second pilot experiment might be misleading. First of all, there is a possibly confounding factor. Contrast an expression from the vague condition: `the square with few dots' with an expression from the crisp condition: `the square with 5 dots'. One difference is that `few' has the potential for vagueness, whereas `5' is crisp. But another difference is that `few' is verbal while `5' is numerical, in the sense that a number is mentioned explicitly. Since these two differences could not be separated in the experiment, the finding of a vagueness advantage is vulnerable to an alternative interpretation, namely that the crisp--vague difference was really an advantage for the verbal form of the quantifier. In our next experiment (reported below as Experiment 1) we therefore created verbal and numeric versions of each of the vague and crisp instructions so that we could compare the crisp--vague difference and the numeric--verbal difference in the same experiment.

In Experiment 1, reported below, we also addressed another issue with the second pilot experiment: participants chose one of two dot arrays --- therefore the `vague' quantifiers (few and many) uniquely identified one square. Recall our definition of vague --- ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise''. The quantifiers `few' and `many' might not have realised their potential for vagueness in the context of a choice between (only) two alternatives --- where there is no borderline alternative that could result in few and many being ``not well-defined''. Furthermore, using definite articles in the instructions may have reinforced the participant's impression that only one choice counted as correct, and this impression could have been reinforced by the use of error feedback.

In the experiments we report below (Experiments 1, 2, and 3) we used what we learned from the pilot experiments to design situations that we believe did address the difference between crisp and vague instructions while taking account of the most important alternative explanations of this crisp--vague difference. The complete data and analysis for the experiments in this paper are available at \texttt{https://mjgreen.github.io/vagueness}.

\section{Experiment 1: Separating vagueness from instruction format}
To find out what happens when words are used in a context where their potential for vagueness comes to the fore, Experiment 1 used three arrays (rather than two) so that the vague description had more than one possible referent, used indefinite articles to avoid the impression that only one response counted as correct, and was carried out without error feedback. An indication that the potential for vagueness was indeed realised in Experiment 1 is that the borderline response was chosen fairly often: 16\% of the time.

In Experiment 1, an item was an instruction followed by a set of three dot arrays defined by a triple of numbers, representing the number of dots in the left, middle, and right arrays. We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). Each set of arrays comprised three arrays (instead of two as in \citet{green2013utility}); the array representing the central number was always presented in the middle of the three; there were two flanking arrays where one had fewer dots than the central array and the other had more, and these flanking arrays appeared equally often on the left and right of the central array.

Table~\ref{instructionsC-exp-1} gives the full set of stimuli and associated instructions. The way in which borderline responses were construed is as follows, using as an example the array (6:15:24) and instructions that identified the smaller flanking array (6). 6 was classified as the expected response. 15 was classified as the borderline response. 24 was classified as the extreme response.

\begin{itemize}
	\item In the vague numerical condition the instruction was ``Choose a square with about 10 dots'' --- none of the arrays contained exactly 10 dots, but 10 is closer to 6 than it is to 15, making 6 a better response to that instruction, 15 a borderline response, and 24 an extreme response.
	\item In the vague verbal condition we used ``Choose a square with few dot''. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
	\item In the crisp numerical condition we used ``Choose the square with 6 dots''. The smaller flanking array always contained exactly the specified number of dots. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
	\item For crisp verbal, we used ``Choose the square with the fewest dots''. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
\end{itemize}

On each trial, first the referring expression that constituted the instruction for that trial was displayed (e.g., ``Choose a square with about 10 dots''). Participants then pressed a key to indicate that they had read the instruction. The instruction remained on screen, and after 1000 ms, the arrays appeared. An example stimulus is given in Figure~\ref{Experiment1and2examplestimulus}. Response time was measured from the presentation of the arrays until the keypress indicating the participant's choice. The trial would timeout after 60 seconds if there was no response. In this experiment, no feedback was given. This was because, in the vague conditions, we did not regard any response as ``correct'' or ``incorrect'', but instead as ``expected response''; ``borderline response''; and ``extreme response'', and we did not want to draw participants' attention to this distinction explicitly. The participant's choice was recorded for analysis.

\begin{table}[htbp]
\centering
\caption{Experiment 1 instructions arranged by condition. The instructions given in the table started with ``Choose \ldots''}
\label{instructionsC-exp-1}
\begin{tabular}{cccll}
\hline\noalign{\smallskip}
Item & Quantity & Number & Crisp & Vague\\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 & Small & Numeric & the square with 6 dots & a square with about 10 dots\\
06:15:24 & Small & Verbal & the square with the fewest dots & a square with few dots\\
06:15:24 & Large & Numeric & the square with 24 dots & a square with about 20 dots\\
06:15:24 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 & Small & Numeric & the square with 16 dots & a square with about 20 dots\\
16:25:34 & Small & Verbal & the square with the fewest dots & a square with few dots\\
16:25:34 & Large & Numeric & the square with 34 dots & a square with about 30 dots\\
16:25:34 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 & Small & Numeric & the square with 26 dots & a square with about 30 dots\\
26:35:44 & Small & Verbal & the square with the fewest dots & a square with few dots\\
26:35:44 & Large & Numeric & the square with 44 dots & a square with about 40 dots\\
26:35:44 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 & Small & Numeric & the square with 36 dots & a square with about 40 dots\\
36:45:54 & Small & Verbal & the square with the fewest dots & a square with few dots\\
36:45:54 & Large & Numeric & the square with 54 dots & a square with about 50 dots\\
36:45:54 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

\begin{figure}[htbp]
\centering
\includegraphics[width=.75\textwidth]{figures/Ce1-example-screenshot-border}
\caption{Experiment 1 and 2 example stimulus}
\label{Experiment1and2examplestimulus}
\end{figure}

\subsection{Hypotheses (Experiment 1)}
We formulated the following hypotheses for Experiment 1:
\begin{description}
	\item [Hypothesis 1 (Crisp/Vague RT)] Vague instructions should result in faster responses than crisp instructions; and this pattern should hold when the model is restricted to numeric-only data and when it is restricted to verbal-only data.
	\item [Hypothesis 2 (Numeric/Verbal RT)] There should be no real difference between responses to numeric instructions and verbal instructions (based on our interpretation of the experiment in \citet{green2013utility}, where we thought that vague instructions alone were driving the advantage for instructions that were both vague, and also in verbal format).
	\item [Hypothesis 3 (Item RT)] Responses should take longer as the number of dots in the display grows larger (i.e., as the levels of \emph{Item} increase).
	\item [Hypothesis 4 (Response Type)] Vague instructions should lead to more borderline responses than crisp instructions.
\end{description}

\subsection{Results (Experiment 1)}
\paragraph{\textbf{Response times}}
30 participants were recruited. Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. The distribution of remaining response times was skewed with many long responses. These remaining response times were log-transformed, which reduced this skew so that their distribution more closely approximated a normal distribution. Condition means for response times are given in Figure~\ref{resultsC-exp-1-RT}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/Ce1-rtplot-1.pdf}
\caption{Experiment 1 results: Mean response times by condition and item}
\label{resultsC-exp-1-RT}
\end{figure}

A linear mixed model was constructed for the (log-transformed) response times, with sum-coded vagueness, instruction format, and their interaction, and item, as fixed effects, and per-participant intercepts and slopes for sum-coded vagueness, instruction format, and their interaction as random effects.

\begin{description}
	\item [Test of Hypothesis 1 (Crisp/Vague RT)] Vague instructions actually led to significantly slower responses than crisp instructions, against Hypothesis 1 ($\beta=0.058$, $se=0.013$, $t=4.55$, $p<0.001$). When the model was restricted to numeric-only instructions, vague instructions still led to significantly slower responses than crisp instructions ($\beta=0.093$, $se=0.021$, $t=4.51$, $p<0.001$). When the model was restricted to verbal-only instructions, vague instructions tended to slow responses, but not significantly ($\beta=0.024$, $se=0.016$, $t=1.46$, $p=0.155$).
	\item [Test of Hypothesis 2 (Numeric/Verbal RT)] There was actually a significant difference between numeric and verbal instructions, with numeric instructions leading to longer responses than verbal instructions, against Hypothesis 2 ($\beta0.265=$, $se=0.072$, $t=5.08$, $p<0.001$).
	\item [Test of Hypothesis 3 (Item RT)] Responses took longer as the levels of \emph{Item} increased, supporting Hypothesis 3 ($\beta=0.120$, $se=0.017$, $t=7.11$, $p<0.001$).
\end{description}

However, given that the response time plot in Figure~\ref{resultsC-exp-1-RT} shows that responses to 6:15:24 in the `crisp numeric' instructions condition were extremely fast relative to the `vague numeric' instructions to 6:15:24, the effects in the model of the full dataset could be driven by this difference. A clearer picture of the effects of interest might be obtained by removing the 6:15:24 level of Item from the data set, and fitting the model to this restricted data. Doing this did not affect the direction of the effects in the full dataset, but whereas the effects were significant in the full dataset, they were not significant in the restricted dataset. Full details of the analysis of the restricted dataset are available in the online materials.

\paragraph{\textbf{Borderline cases}}

A generalized linear mixed model \citep{jaeger2008categorical} was fit to the data for the distribution of responses indicating the borderline response, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant, as well as per-participant intercepts as random effects. The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Figure~\ref{resultsC-exp-1-BL}. Participants chose the borderline square on $16.6\%$ of trials overall.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/Ce1-blBarChart-1}
\caption{Experiment 1 results: Counts of borderline case responses by condition.}
\label{resultsC-exp-1-BL}
\end{figure}

\begin{description}
	\item [Test of Hypothesis 4 (Response Type)] Participants were significantly more likely to choose the borderline option for vague instructions than for crisp instructions (21.9\% vs 11.3\%: $\beta=0.62$, $se=0.22$, $z=2.8$, $p=0.0059$). Participants were also significantly more likely to choose the borderline square when the instruction used the numerical format rather than the verbal format (30.1\% vs 3.0\%: $\beta=-3.35$, $se=0.23$, $z=-14.6$, $p<0.0001$).
\end{description}

\subsection{Discussion (Experiment 1)}

This experiment tested whether vague instructions would result in faster responses than crisp instructions, when borderline cases were present. Faster responses for vague instructions were found in pilot experiment B, but there were no borderline cases in that experiment.

In this experiment we found in contrast that vague instructions resulted in slower responses than crisp instructions: a difference that was significant when considering the full data (112ms), but which was not significant after removing the smallest arrays from the analysis, which had a pattern opposite to the main trends in the rest of the data.

We also found that the effect of instruction format was significant, with numerical format slowing responses by 689 ms on average, such that the disadvantage of numerical format overwhelmed the contribution of vagueness. The verbal vague condition still yielded faster responses than the numerical crisp condition, so the pattern from pilot experiment B was reproduced, but in the light of the evidence from this experiment (Experiment 1), in the presence of borderline cases, the advantage that was ascribed to vagueness before now looks more like an advantage of verbal instruction format.

However, once again there is a possibly confounding factor. Observe that, in Experiment 1, instruction format (i.e., the difference between numeric and verbal) went hand in hand with what might be called the (human) `selection algorithm': To see this, consider the task of selecting the dot array that contains `few dots': to do this, it suffices to \emph{compare} the three arrays and select the one that contains the fewest elements.  To select the dot array that contains `16 dots' seems to require the participant to estimate, and then \emph{match}, the cardinality of (at least) one dot array to 16, a process which could plausibly take longer, independently of vagueness. Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use \emph{comparison} rather than having to resort to \emph{matching}.

In the next two experiments we pitted the comparison algorithm and matching algorithm selection tasks against each other while controlling vagueness and instruction format. In Experiment 2 we restricted all the instructions to numeric quantifiers while factorially manipulating vagueness and selection task. In Experiment 3 we ensured that all instructions used verbal quantifiers, while also factorially manipulating `vagueness' and `selection task'. This allowed us to distinguish between the predictions of the selection task account and the instruction format account.

\section{Experiment 2: Focus on instructions that contain numerals}
The main aim of Experiment 2 was to see whether vagueness would exert beneficial effects when all conditions used numerals in the instructions, and when there were vague and crisp versions of the instructions for both comparison and matching strategies. The main changes from Experiment 1 were that the human selection task was explicitly controlled (i.e., whether the task amounted to matching or comparison), and that all conditions were constrained to mention a number. We used the same arrays as in Experiment 1 (an example stimulus is given in Figure~\ref{Experiment1and2examplestimulus}). We used a 2 x 2 factorial manipulation of vagueness and selection task (see Table~\ref{Instructions for e2}). On each trial an instruction was presented: participants pressed a key to dismiss the instruction, at which time the dot arrays were presented until the participant responded, and the response time and choice were recorded. Table~\ref{Instructions for e2} shows the instructions for each condition. Note the difference between ``fewer than 20'' and ``far fewer than 20'': whereas the former cannot have borderline cases (i.e., for each number it is clear whether the number is smaller than 20 or not), the latter can.

\begin{table}
\centering
\caption{Experiment 2: Instructions arranged by condition. The instructions given in the table started with ``Choose a square with \ldots''}
\label{Instructions for e2}
\begin{tabular}{cccll}
\hline\noalign{\smallskip}
Item & Quantity & Selection & Crisp & Vague \\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 & Small & Comparison & fewer than 20 dots & far fewer than 20 dots \\
06:15:24 & Small & Matching & 6 dots & about 10 dots \\
06:15:24 & Large & Comparison & more than 10 dots & far more than 10 dots \\
06:15:24 & Large & Matching & 24 dots & about 20 dots \\
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 & Small & Comparison & fewer than 30 dots & far fewer than 30 dots \\
16:25:34 & Small & Matching & 16 dots & about 20 dots \\
16:25:34 & Large & Comparison & more than 20 dots & far more than 20 dots \\
16:25:34 & Large & Matching & 34 dots & about 30 dots \\
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 & Small & Comparison & fewer than 40 dots & far fewer than 40 dots \\
26:35:44 & Small & Matching & 26 dots & about 30 dots \\
26:35:44 & Large & Comparison & more than 30 dots & far more than 30 dots \\
26:35:44 & Large & Matching & 44 dots & about 40 dots \\
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 & Small & Comparison & fewer than 50 dots & far fewer than 50 dots \\
36:45:54 & Small & Matching & 36 dots & about 40 dots \\
36:45:54 & Large & Comparison & more than 40 dots & far more than 40 dots \\
36:45:54 & Large & Matching & 54 dots & about 50 dots \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

\subsection{Hypotheses (Experiment 2)}

For Experiment 2, we formulated the following hypotheses:

\begin{description}
	\item [Hypothesis 1 (Crisp/Vague RT)] Vague instructions should result in faster responses than crisp instructions.
	\item [Hypothesis 2 (Comparison/Matching RT)] Instructions that allow comparison should result in faster responses than instructions that necessitate matching.
	\item [Hypothesis 3 (Interaction)] The vagueness effect should differ according to whether the selection task is comparison or matching.
\end{description}

\subsection{Results (Experiment 2)}

38 participants were recruited. Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 204 trials (2.8\% of the trials). The distribution of remaining response times was skewed with many long responses. These remaining response times were log-transformed, which reduced this skew so that their distribution more closely approximated a normal distribution. Condition means for response times are plotted in Figure~\ref{resultsD-exp-2}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/De2-rtplot-1.pdf}
\caption{Mean response times by condition for Experiment 2 where all instructions were numeric}
\label{resultsD-exp-2}
\end{figure}

A linear mixed model was constructed for the logged response times, with sum-coded vagueness, selection task, and their interaction, and item as fixed effects, and per-participant intercepts and slopes for sum-coded vagueness, selection task, and their interaction, and item for random effects.
